A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-05-29 introduction.tex
Introduction

\subsection{An intuition for our method}

In this section we will introduce an intuition for how a string constraint problem in an automata-based solver like \OstrichPlus{} is translated to a Parikh automata intersection problem and solved efficiently using \Calculus{}. First, by a Parikh automaton we mean a standard Non-deterministic finite automaton (NFA) with a set of Parikh registers that are incremented (or decremented) at each transition. In the automata of \cref{fig:examples}, we will use the notation $a / \begin{bmatrix} x_+ \end{bmatrix}$ to mean that a transition reads an input character $a$ and increments register $x$ by one. We will omit any zero-valued increments and assume that all registers are scoped to their automaton.

Furthermore, we use $\CountOf{c}(A)$ to refer to the number of letters $c$ in $A$, and allow this expression to be used both in the input language of \cref{ex:string-constraints} and in the langauge of constraints that we translate to. Where $A$ is apparent, we will omit it.

@@

Each transition is associated with a fresh variable detailing how many times it is used

Counters are the sum of their increments times number of times the transition was used

TODO: rewrite the input and log it through CATRA!

\begin{figure}[h]
    \centering 
\begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b}
    \caption{The automaton $B$, where the Parikh registers count the number of occurrences of each letter.}\label{fig:aut_b}
      %\Description[SHORT]{LONG}
  \end{subfigure}

  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a}
    \caption{The automaton $A'$, whose Parikh registers count both the number of letters and the start ($l$) and length ($r$) of the substring.}\label{fig:aut_a}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The collection of automata we use as running examples}\label{fig:examples}
\end{figure}

>>>>

\subsection{An Intuition for Our Approach}\label{sec:intuition}

\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a}
    \caption{The automaton $A'$, whose Parikh registers count both the number of letters and the start ($l$) and length ($r$) of the substring matching $\Regex{c\RegexOr{}dd}$.}\label{fig:aut_a}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b}
    \caption{The automaton $B$, where the Parikh registers count the number of occurrences of each letter.}\label{fig:aut_b}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The collection of automata we use as running examples}\label{fig:examples}
\end{figure}

In this section we will introduce an intuition for how a string constraint problem in an automata-based solver like \OstrichPlus{} is translated to a Parikh automata intersection problem and solved efficiently using \Calculus{}. First, by a Parikh automaton we mean a standard Non-deterministic finite automaton (NFA) with a set of Parikh registers that are incremented (or decremented) at each transition. In the automata of \cref{fig:examples}, we will use the notation $a / \begin{bmatrix} x_+ \end{bmatrix}$ to mean that a transition reads an input character $a$ and increments register $x$ by one. We will omit any zero-valued increments and assume that all registers are scoped to their automaton. The increments are in reality a vector (hence the notation), but as it is mostly sparse here, we instead use the symbolic notation $\begin{bmatrix} x_{1+} \end{bmatrix}$ rather than the more cumbersome $\begin{bmatrix} 0 \\ 1 \\ 0 \\  \vdots \\ 0 \end{bmatrix}$.

Furthermore, we use $\CountOf{c}(A)$ to refer to the number of letters $c$ in $\Language(A)$, and allow this expression to be used both in linear constraints of the input language of \cref{ex:string-constraints} and in the langauge of constraints that we translate to. Where the automaton $A$ is apparent, we will omit it.

@@

Our approach to the computation of the Parikh image (and therefore the solution to the intersection problem posed by \cref{ex:string-constraints}) consists of two applications of laziness; late computation of products of automata, and lazy enforcement of connectivity constraints. Both are in contrast to the eager approach to finding the Parikh image described in~\cite{generate-parikh-image}.

Starting with an interactive theorem prover and the goal of solving the constraint $\CountOf{\Char{a}}(B \times A') > \CountOf{\Char{b}}(B \times A')$ we associate each transition of $A'$ and $B$ with a fresh integer variable, $\TransitionVar_\Transition$, identically to \cite{generate-parikh-image}. These variables represent how many times each transition is taken. Hence, the final counter value for each automaton is the element-wise sum:
\begin{equation}
\begin{bmatrix} \CountOf{\Char{a}} \\ \CountOf{\Char{b}} \\ \CountOf{\Char{c}} \\ \CountOf{\Char{d}} \end{bmatrix} = \sum_\Transition \TransitionVar_\Transition \cdot \IncrementVec_\Transition \text{ where $\IncrementVec_\Transition$ are the increments of transition $\Transition$}.
\end{equation}

We start by adding linear constraints requiring transition variables to represent a flow through either automata by requiring that the number of incoming transitions is equal to the number of outgoing ones, and use the automated theorem prover to annotate each transition $\Transition$ with a symbolic representation of its associated integer variable:
\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a_annotated}
    \caption{Note: unused self-loops at $I_{A'}$ and $F_{A'}$ have been pruned for readability.}\label{fig:aut_a_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b_annotated}
    \caption{}\label{fig:aut_b_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The starting automata, without their counter increments and with their transition variables in symbolic form.}\label{fig:propagated}
\end{figure}

Note that several transitions are already marked as unuseable (have a transition value of $0$). This includes any transition for the letter $\Char{d}$, as all counter increments for that variable in $B$ is zero and hence it must also be zero in the product of both automata. By removing these transitions to obtain new automata and computing their product, we obtain the following:

\begin{figure}[h]
    \centering 
    \includegraphics[scale=\autscale]{ab}
  \caption{An automaton encoding the constraint $\CountOf{\Char{a}}(B \times A') > \CountOf{\Char{b}}(B \times A')$.}\label{fig:product}
      %\Description[SHORT]{LONG}
\end{figure}

By putting off computing the product $B \times A'$ until after performing linear reasoning on the number of times each transition must be taken, we have constructed a much smaller product than we would have originally. Moreover, since this automaton is a stick, all runs through it can be described by the flow-preservation equations described above without the need for costly connectivity-enforcing constraints as in~\cite{generate-parikh-image}. This illustrates the advantage of our second laziness: deferred enforcement of connectivity constraints.

### 2023-05-29 main.tex
Add new examples
More work on introduction
Introduction

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

%% Appendix
\appendix
\section{Appendix}
\subsection{Per-instance comparison between backends}
\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-baseline-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:baseline}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and baseline]{The plot shows most dots at the bottom of the plot, with some plots collected at the right-hand side.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-nuxmv-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:nuxmv}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and \Nuxmv{}]{The plot shows flames of dots between the bottom and top, with a line of dots near the right-hand side corresponding to timeouts.}
  \end{subfigure}
\end{figure}
\subsection{Example instance for \Catra{}}
\begin{lstlisting}[caption={An example input file for \Catra{} for the problem introduced in \cref{sec:motivation}, illustrating every major syntax element. From beginning to end: synchronised (product) automata using the keyword \texttt{synchronised} (automata A and B), labels (except those with ranges), register increments, and constraints on the final values of their counters.}, label=lst:input-example]
  // So far we only support integer counters. Note that we have individual
  // counters for each automaton to avoid surprises for product construction.
  counter int l_a, l_b, l_c, r_c;

  synchronised {
  automaton aca_or_bc {
    init S;

    // We use ASCII values here, this is for lowercase a
    S -> A [97] { l_a += 1 };
    S -> B [98] { l_b += 1 };

    B -> S  [98] { l_b += 1 };
    B -> final [99] { l_c += 1 };

    A -> A [99] { l_c += 1 };
    A -> final [97] { l_a += 1 };

    accepting final;
  };

  automaton something_c_something {
      init S;

      // Special short-hand value for the whole Unicode alphabet.
      S -> S [any] ;
      S -> F [99] { r_c += 1 };

      F -> F [any];

      accepting F;
  };
  };

  // Constrain the number of a:s to be larger than the number of c:s. Since the
  // automata are synchronised on every transition and counters are guaranteed
  // to be consistent these constraints are sufficient without also constraining
  // r_c.
  constraint l_a > l_c;
\end{lstlisting}

>>>>

\usepackage{enumitem}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}{Example}[section]

\newlist{constraints}{enumerate}{3}
\setlist[constraints,1]{%
    label=\sffamily{(\roman*):},
    ref=\normalfont{Constraint (\roman*)},
    wide,itemsep=0pt,topsep=0pt
}    
\crefname{constraintsi}{}{}
\Crefname{constraintsi}{}{}

@@

%% Appendix
\appendix
\section{Appendix}
\subsection{Per-instance comparison between backends}
\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-baseline-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:baseline}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and baseline]{The plot shows most dots at the bottom of the plot, with some plots collected at the right-hand side.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-nuxmv-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:nuxmv}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and \Nuxmv{}]{The plot shows flames of dots between the bottom and top, with a line of dots near the right-hand side corresponding to timeouts.}
  \end{subfigure}
\end{figure}
\subsection{Example instance for \Catra{}}
\begin{lstlisting}[caption={An example input file for \Catra{} for \cref{ex:string-constraints} in \cref{sec:intuition}, illustrating every major syntax element. From beginning to end: synchronised (product) automata using the keyword \texttt{synchronised} (automata A and B), labels (except those with ranges), register increments, and constraints on the final values of their counters.}, label=lst:input-example]
\input{example.par}
\end{lstlisting}
